cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-13 paper-cardinalR.Rmd
Updating branching section

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology [@trapnell2014]. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (`n`), and number of branches (`k`). Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

Branches $1$ and $2$ are initialized with fixed slopes and intercepts, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

@@

where $s_i$ is a scale factor controlling the curvature of branch $i$. For the first few initial branches, $s_i$ can be fixed (e.g., $s_1 = 1, s_2 = 2$), while for subsequent branches it is sampled from a predefined set, such as $s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}$, to create variability in curvature.

@@

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) and `gen_orgcurvybranches(n, p, k)`. Each branch is embedded in a unique or repeated $2\text{-}D$ subspace of the $p\text{-}D$ space. When `allow_share = TRUE`, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible $\binom{p}{2}$ combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow

@@

where $a_i, b_i$ define the range of the first coordinate for branch $i$, and $\epsilon$ is Gaussian noise added to introduce variability. The scale factor $s_i$ controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first $\binom{p}{2}$ branches, $s_i = 1$; for additional branches when $k > \binom{p}{2}$, $s_i$ is randomly drawn from the set $\{1, 1.5, 2, \dots, 8\}$.

@@

```{r data-clusteredspheres}
clusteredspheres <- gen_clusteredspheres(n = c(1000, 100), k = 3, r = c(15, 3),
                                         loc = 10 / sqrt(3)) |>
  dplyr::select(-cluster)
```

>>>>

A branching structure (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology [@trapnell2014]. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, total number of points (`n`) generated across all branches, with points allocated approximately evenly among branches, and number of branches (`k`). Although these functions can generate multiple branches, they do not produce a formal *multicluster* dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

The jitter term is sampled from a one-sided uniform distribution to introduce directional variability without altering branch orientation.

Branches $1$ and $2$ are initialized with fixed slopes and starting points, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

@@

where $s_i$ is a scale factor controlling the curvature of branch $i$. For the first few initial branches, $s_i$ can be fixed (e.g., $s_1 = 1, s_2 = 2$) to establish reference shapes, while for subsequent branches it is sampled from a predefined set, such as $s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}$, to control curvature magnitude and direction.

@@

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch)"} else { "Figure \\@ref(fig:branch-proj)"})`) and `gen_orgcurvybranches(n, p, k)`. For branch 
$i$, the active coordinate pair $(i_1, i_2)$ indexes the selected $2\text{-}D$ subspace. When `allow_share = TRUE`, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible $\binom{p}{2}$ combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow

@@

where $a_i, b_i$ define the range of the first coordinate for branch $i$, and $\epsilon$ is Gaussian noise added to introduce smooth variability in $p\text{-}D$ without sharp boundaries. The scale factor $s_i$ controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first $\binom{p}{2}$ branches, $s_i = 1$; for additional branches when $k > \binom{p}{2}$, $s_i$ is randomly drawn from the set $\{1, 1.5, 2, \dots, 8\}$.

Across all branching generators, the scale parameter $s_i$ controls the strength of deviation from linearity, determining slope, curvature, or growth rate depending on the branch geometry.

@@

```{r data-clusteredspheres}
clusteredspheres <- gen_clusteredspheres(n_vec = c(1000, 100), k_small = 3, r_vec = c(15, 3),
                                         sep = 10 / sqrt(3)) |>
  dplyr::select(-cluster)
```

### 2025-12-13 paper-cardinalR.tex
Updating branching section

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology \citep{trapnell2014}. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, numbers of points (\texttt{n}), and number of branches (\texttt{k}). Although these functions can generate multiple branches, they do not produce a formal \emph{multicluster} dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

@@

where \(s_i\) is a scale factor controlling the curvature of branch \(i\). For the first few initial branches, \(s_i\) can be fixed (e.g., \(s_1 = 1, s_2 = 2\)), while for subsequent branches it is sampled from a predefined set, such as \(s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}\), to create variability in curvature.

@@

High-dimensional generalizations are provided by \texttt{gen\_orglinearbranches(n,\ p,\ k)} (Figure \ref{fig:branch-proj}) and \texttt{gen\_orgcurvybranches(n,\ p,\ k)}. Each branch is embedded in a unique or repeated \(2\text{-}D\) subspace of the \(p\text{-}D\) space. When \texttt{allow\_share\ =\ TRUE}, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible \(\binom{p}{2}\) combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow

@@

where \(a_i, b_i\) define the range of the first coordinate for branch \(i\), and \(\epsilon\) is Gaussian noise added to introduce variability. The scale factor \(s_i\) controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first \(\binom{p}{2}\) branches, \(s_i = 1\); for additional branches when \(k > \binom{p}{2}\), \(s_i\) is randomly drawn from the set \(\{1, 1.5, 2, \dots, 8\}\).

>>>>

A branching structure (Figure \ref{fig:branch-proj}) captures trajectories that diverge or bifurcate from a common origin, similar to processes such as cell differentiation in biology \citep{trapnell2014}. We introduce a set of data generation functions specifically designed to simulate high-dimensional branching structures with various geometries, total number of points (\texttt{n}) generated across all branches, with points allocated approximately evenly among branches, and number of branches (\texttt{k}). Although these functions can generate multiple branches, they do not produce a formal \emph{multicluster} dataset: the branches form a single connected structure, with multiple visually distinct arms rather than independent clusters.

@@

The jitter term is sampled from a one-sided uniform distribution to introduce directional variability without altering branch orientation.

Branches \(1\) and \(2\) are initialized with fixed slopes and starting points, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

@@

where \(s_i\) is a scale factor controlling the curvature of branch \(i\). For the first few initial branches, \(s_i\) can be fixed (e.g., \(s_1 = 1, s_2 = 2\)) to establish reference shapes, while for subsequent branches it is sampled from a predefined set, such as \(s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}\), to control curvature magnitude and direction.

@@

High-dimensional generalizations are provided by \texttt{gen\_orglinearbranches(n,\ p,\ k)} (Figure \ref{fig:branch-proj}) and \texttt{gen\_orgcurvybranches(n,\ p,\ k)}. For branch
\(i\), the active coordinate pair \((i_1, i_2)\) indexes the selected \(2\text{-}D\) subspace. When \texttt{allow\_share\ =\ TRUE}, multiple branches may share the same subspace; otherwise, subspaces are sampled without replacement until all possible \(\binom{p}{2}\) combinations are exhausted, after which additional branches may repeat subspaces. Linear branches follow

@@

where \(a_i, b_i\) define the range of the first coordinate for branch \(i\), and \(\epsilon\) is Gaussian noise added to introduce smooth variability in \(p\text{-}D\) without sharp boundaries. The scale factor \(s_i\) controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first \(\binom{p}{2}\) branches, \(s_i = 1\); for additional branches when \(k > \binom{p}{2}\), \(s_i\) is randomly drawn from the set \(\{1, 1.5, 2, \dots, 8\}\).

Across all branching generators, the scale parameter \(s_i\) controls the strength of deviation from linearity, determining slope, curvature, or growth rate depending on the branch geometry.
